\documentclass[fleqn]{article}
\usepackage{mystyle}

\title{Programming Project 5: UHF\\
\textit{Unrestricted Hartree-Fock\\(spin-orbital)}}
\author{}
\date{}

\begin{document}

\maketitle

\noindent
\bmit{Spatial AO basis $\{\x_\mu\}$ integrals.}
\begin{align}
\label{space-ao-ints}
&
  \ol{S}_{\mu\nu}
=
  \ip{\x_{\mu}|\x_{\nu}}
&&
  \ol{T}_{\mu\nu}
=
  -\tfrac{1}{2}\ip{\x_{\mu}|\nabla_1^2|\x_{\nu}}
&&
  \ol{V}_{\mu\nu}
=
  \sum_A\ip{\x_{\mu}|\fr{Z_A}{|\bo{r}_1-\bo{R}_A|}|\x_{\nu}}
&&
  (\mu\nu|\rho\si)
=
  \ip{\x_\mu\x_\rho|\fr{1}{|\bo{r}_1-\bo{r}_2|}|\x_\nu\x_\si}
\end{align}


\noindent
\bmit{Spin-AO basis $\{\x_\mu\a\}\cup\{\x_\mu\b\}$ integrals.}
\begin{align}
\label{spin-ao-ints-1}
&
  \bo{S}
=
  \ma{\ol{\bo{S}}&0\\0&\ol{\bo{S}}}
&&
  \bo{T}
=
  \ma{\ol{\bo{T}}&0\\0&\ol{\bo{T}}}
&&
  \bo{V}
=
  \ma{\ol{\bo{V}}&0\\0&\ol{\bo{V}}}
&&
  \bo{G}
=
  \ma{\ol{\bo{G}}&0\\0&\ol{\bo{G}}}
&&
  (\bo{G})_{\mu\nu}
=
  \ma{(\ol{\bo{G}})_{\mu\nu}&0\\0&(\ol{\bo{G}})_{\mu\nu}}
\end{align}
\begin{align}
\label{spin-ao-ints-2}
&
\text{where $\ol{\bo{G}}$ represents $(\mu\nu|\rho\si)$ viewed as a matrix of matrices:}
\ \ 
  \ol{\bo{G}}
=
  \ma{(\ol{\bo{G}})_{\mu\nu}},
\ \ 
  (\ol{\bo{G}})_{\mu\nu}
=
  [((\ol{\bo{G}})_{\mu\nu})_{\rho\si}]
=
  [(\mu\nu|\rho\si)]
\end{align}


\noindent
\bmit{Working equations.}
\begin{align}
\label{fock-matrix}
&&
  f_{\mu\nu}
=&\
  h_{\mu\nu}
+
  v_{\mu\nu}
\sp
  h_{\mu\nu}
=
  T_{\mu\nu}+V_{\mu\nu}
\sp
  v_{\mu\nu}
=
  \sum_{\rho\si}
  \ip{\mu\rho||\nu\si}
  D_{\rho\si}
\sp
  D_{\mu\nu}
=
  \sum_{i=1}^{n_\text{occ}}
  C_{\mu i}^* C_{\nu i}
\\
\label{energy}
&&
  E
=&\
  E_e
+
  V_\mathrm{Nu}
\sp
  E_e
=
  \sum_{\mu\nu}
  (h_{\mu\nu}+\tfrac{1}{2}v_{\mu\nu})D_{\mu\nu}
\\
\label{orthogonalizer}
&&
  \bo{X}
=&\
  \bo{S}^{-\frac{1}{2}}
\sp
  \bo{\tl{f}}
=
  \bo{X}\bo{f}\bo{X}
\sp
  \bo{C}
=
  \bo{X}
  \bo{\tl{C}}
\end{align}

\subsection*{Procedure}

\bmit{Initialization.}
\begin{enumerate}
  \item Read in nuclear repulsion energy ($V_\mathrm{Nu}$) from the Molecule object
  \item Read in spatial AO basis integrals from LibMints (eq. \ref{space-ao-ints})
  \item Determine spin-AO basis integrals from spatial AO basis integrals (eq. \ref{spin-ao-ints-1} and \ref{spin-ao-ints-2})
  \item Form orthogonalizer $\bo{X}$ (eq. \ref{orthogonalizer})
\end{enumerate}

\noindent
\bmit{Iteration to self-consistency.}
\begin{enumerate}
  \item Set $D_{\mu\nu}=0$ as starting guess\footnote{This is the ``core'' guess, because $F_{\mu\nu}$ becomes the core Hamiltonian $h_{\mu\nu}$}
  \item\label{loop} Build Fock matrix (eq.~\ref{fock-matrix})
  \item Transform $\bo{f}\mapsto\bo{\tl{f}}$ to orthogonalized AO basis (eq.~\ref{orthogonalizer})
  \item Diagonalize $\bo{\tl{f}}$, yielding orbital energies $\ev_p$ and MO coefficients $\tl{C}_{\mu p}$
  \item Backtransform $\bo{\tl{C}}\mapsto\bo{C}$ to original AO basis (eq.~\ref{orthogonalizer})
  \item Build density matrix $\bo{D}$ (eq.~\ref{fock-matrix})
  \item If the energy (eq.~\ref{energy}) is not converged, return to step \ref{loop}.
\end{enumerate}

\end{document}
